~\Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores

### 2024-09-30 manuscript/paper.tex
Update manuscript with posterior and Kendall's Tau

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

@@

The lineage II fossil data consist of 814 specimens of unknown sex
sampled across 18 time periods spaced \textasciitilde1000 years apart.
Figure 1 shows the sample size for each of the 18 samples. There are at
least 22 specimens in each sample with a high of 67 specimens in sample
7.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & \texttt{stl} & Distance from anterior tip of
premaxilla to posterior end of last vertebra (hypural plate) \\
Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
to the anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
including the first vertebra contacting an anal fin pterygiophore
(Aguirre et al.~2014) \\
Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
to but excluding the pterygiophore under the third dorsal spine, which
is immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
pelvic spine above its articulation with the pelvic girdle to its distal
tip \\
Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
posterior tips along midline. If vestigial, the sum of longest
anterior-posterior axis for the vestiges \\
Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
ventral tip of the cleithrum on the anterior margin of the shoulder
girdle (i.e., cleithrum) \\
Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
premaxilla to the distal tip of the ascending process of the
premaxilla \\
Dorsal Spine & \texttt{Ds\#.sc\#\ =\ 1,2,or\ 3} & Length from the base
of a dorsal spine above the pterygiophore to its distal tip along the
anterior edge \\
Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
posterior tips of the pterygiophore immediately preceding the 3rd dorsal
spine (when present) \\
\bottomrule()
\end{longtable}

@@

Briefly, We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Imputations were performed using the MICE (Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) algorithm and
implemented in R (Team (\protect\hyperlink{ref-R2022language}{2022}))
with M = 100 imputations. After imputing the sex, we fit a modified
Orenstein-Uhlenbeck (OU; Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})) model to each imputed data
set using a Bayesian framework to \{THIS NEEDS MORE\}. We then
quantified sexual dimorphism as the \{THIS NEEDS MORE; MAYBE WE CAN
WRITE TOGETHER\}. We describe the details of the imputation model and OU
modeling next.

\hypertarget{sec:models}{%
\section{Models}\label{sec:models}}

\hypertarget{imputation-model}{%
\subsection{Imputation Model}\label{imputation-model}}

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate sex of the stickleback fish and
\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) phenotypes of interest. Because the sex of the fossilized
stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively.

\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}

We impute missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if users intend on performing Bayesian analysis
after imputation. Therefore, we imputed \(M = 100\) complet datasets.

\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}

We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than
usingRubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
distributions in our Bayesian analysis.

\hypertarget{bayesian-ou-modeling}{%
\subsection{Bayesian OU modeling}\label{bayesian-ou-modeling}}

For Bayesian modeling, we dropped the extant data and focused on
modeling the evolution of sexual dimorphism through time in the fossils.

\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
our model; connecting to the previous section, \(W_{ij}\) and
\(\boldsymbol{Y}_{ij}\) can be interpreted as \(W_{fossil,ij}\) and
\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
the fish, \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits of
stickleback fish will have some correlation with its standard length
\textbf{(CITATION)}. We account for this by adding an additional
parameter, \(\gamma_k\), into our model. More specifically, if
\(Y_{k,ij}\) is a continuous trait, then \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If \(Y_{k,ij}\) is a discrete trait, the conventional method of
modelling this data is by fitting a Poisson distribution. However, the
Poisson distribution assumes that the mean and variance of \(Y_{k,ij}\)
are equal (\(E(Y_{k,ij}) = \text{Var}(Y_{k,ij})\)), while the empirical
fossil data may have variances that are smaller than their respective
means. To combat this issue, we propose to fit the discrete traits to a
generalized Poisson model as defined in
(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align} where \(E(X) = \lambda\) and
\(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\). If \(\alpha > 0\), then the
variance is greater than the mean, called overdispersion; if
\(\alpha < 0\), then the variance is greater than the mean, called
underdispersion; if \(\alpha = 0\), then the model degenerates to a
Poisson distribution.

We will use the generalized Poisson to model all of the discrete traits
in our dataset. In addition, we assume that the discrete traits
abdominal vertebrae (mav) and caudal vertebrae (mcv) also have
correlation with the standard length of the fish (\textbf{CITATION}). If
\(Y_{k,ij}\) is one of the above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we also assume the above model except we
set \(\gamma_k = 0\) because of the assumption of no correlation between
the standard length and these traits. More specifically, we assume
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

@@

The lineage II fossil data consist of 814 specimens of unknown sex
sampled across the 18 K series samples. Figure 1 shows the sample size
for each of the 18 samples. There are at least 22 specimens in each
sample with a high of 67 specimens in sample 7.

THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Standard Length & \texttt{stl} & Distance from anterior tip of
premaxilla to posterior end of last vertebra (hypural plate) \\
Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
to the anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
including the first vertebra contacting an anal fin pterygiophore
(Aguirre et al.~2014) \\
Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
to but excluding the pterygiophore under the third dorsal spine, which
is immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
pelvic spine above its articulation with the pelvic girdle to its distal
tip \\
Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
posterior tips along midline. If vestigial, the sum of longest
anterior-posterior axis for the vestiges \\
Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
ventral tip of the cleithrum on the anterior margin of the shoulder
girdle (i.e., cleithrum) \\
Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
premaxilla to the distal tip of the ascending process of the
premaxilla \\
Dorsal Spine & \texttt{Ds\#.sc\#\ =\ 1,2,or\ 3} & Length from the base
of a dorsal spine above the pterygiophore to its distal tip along the
anterior edge \\
Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
posterior tips of the pterygiophore immediately preceding the 3rd dorsal
spine (when present) \\
\end{longtable}

@@

Briefly, We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Let \(\boldsymbol{W}\) be an
\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
the stickleback fish and \(\boldsymbol{Y}\) be an
\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
interest. Because the sex of the fossilized stickleback fish is
unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively. We imputeD
missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching, implemented in R (CITATION).
Traditionally, the choice for the number of completed dats sets is a
relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
larger number of imputed data sets if users intend on performing
Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
complete datasets.

\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than using
Rubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
distributions in our Bayesian analysis.

\hypertarget{modeling-trait-means-through-time-for-males-and-females}{%
\subsection{Modeling trait means through time for males and
females}\label{modeling-trait-means-through-time-for-males-and-females}}

The goal of our analysis was to estimate fossil trait means through
time, by sex, while incorporating uncertainty in the imputed data,
especially uncertainty sex classification. We took a Bayesian approach
because imputation is the process of sampling from a posterior
predictive distribution of missing data, given observed data.
\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH. -YS}

\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
To estimate fossil sample means through time, we first dropped the
extant individuals from our imputed data sets. Then, for a given imputed
dataset, we let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). In addition, we denote \(Y_{K,ij}\), the last
variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of each
fish individual, a measure of body size. \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits will have
some correlation with standard length \textbf{(CITATION)}. We account
for this by adding an additional parameter, \(\gamma_k\), If
\(Y_{k,ij}\) is a continuous trait, then \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If \(Y_{k,ij}\) is a discrete trait, the conventional modelling method
is to fit a Poisson distribution. However, because the empirical fossil
data violate the assumption that the Poisson variance is equal to its
mean, we fit discrete traits to a generalized Poisson model as defined
in
(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align} where GP denotes a generalized Poisson distribution with
\(E(X) = \lambda\) and \(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\).

In addition, we allowed that two discrete traits, abdominal vertebrae
number (mav) and caudal vertebrae number (mcv) could also with standard
length (\textbf{CITATION}). This is because these are vertebral counts
that run in parallel with the standard length measurement, along the
anterior-posterior access of the fish; these vertebrae thus contribute
to the length of the fish directly, and may evolve in a correlated
fashion. If \(Y_{k,ij}\) is one of the above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we set \(\gamma_k = 0\) because we expect
that trait counts are set during early ontogeny, and should not change
with size in adult fish. Thus, \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align} We point out that, for the discrete traits, the means are
represented by \(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for
ease of modeling.

In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
model the time-\(t_i\), sample-specific trait mean \(k\) for females and
males, respectively. To account for the possibility that there is a time
-dependent trend in \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\), we further
set \begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\), where \(\beta_{0,kg}\) and
\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
sex, and \(u_{k,gt_i}\) is the corresponding residual. To account for
potential correlations between the residuals for a given trait \(k\) and
sex \(g\), we fit an Ornstein-Uhlenback (OU) process (Uhlenbeck and
Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More specifically,
we defined \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the change in
\(u_k,gt\) for a given trait \(k\) and sex \(g\) over a miniscule time
period \(dt\). The OU process is defined as \begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align} where \(\kappa_k\) is a parameter associated with the
correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
the standard deviation of the OU process, and \(W_t\) is a standard
Brownian motion. As shown in Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930}), the closed form solution for
the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
constant. This unknown constant can therefore be considered a latent
variable, further justifying our Bayesian approach, which better
incorporates prior information and uncertainty from missing data
imputation in the estimate of the variable.

We discuss the procedure for estimating this value below.
\textcolor{red}{DID THIS GET DISCUSSED?-YS}

Priors: To aid in the sampling procedure for the discrete phenotypes, we
first had to transform the parameters of the generalized Poisson
distribution onto the real number line in order to ensure our algorithm
can properly sample from the posterior distribution of interest using
STAN software. This was done by declaring
\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
where
\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).

We chose the following priors, designing them to be weakly informative
such that the posterior distributions could be influenced by the data,
if warranted. For \(k = 1,\cdots,K\),
